---
title: "Intro to R - Spring 2023"
author: "Angela Rexwinkle"
date: "2023-01-26"
output: html_document
---

Many portions of this introduction were used from other
online sources. Some of these include:
Tutorial from LISA at UC Boulder<https://osf.io/6jb9t>
R Cheat Sheet <https://iqss.github.io/dss-workshops/R/Rintro/base-r-cheat-sheet.pdf>
R Primer dashboard - practice!
<https://posit.cloud/learn/primers>

Assessment 
<https://jayholster.shinyapps.io/RLevel0Assessment/>


```{r}
x <- 10 # This is an example of a comment.
```


Follow prompts to install.

```{r install.packages, error = TRUE}
install.packages("tidyverse")
```

```{r}
library(tidyverse)

```


You will need to run the code chunk in order to process the change in working directory. To run a code chunk, you can select the code you want to run and hit command and enter simultaneously on Mac, or control and enter on a Windows machine. Alternatively, you can click the green play button in the top right corner of the code chunk to run the entire cell.

```{r workingdirectory, eval = F}
setwd('/Users/username/Desktop/R/Short Courses')
```



```{r number}
7
```


```{r mathematical}
7+2
```


```{r multiple objects}
7+2-10*40
```


```{r sum function}
sum(1, 10)
```


This won't work. Why? 

```{r sum function error, error = T}
sum(1 10)
```

You can read this code aloud saying "sum all
numbers from 1 to 10."

```{r sum sequence}
sum(1:10)
```

you might
encounter a new function which you are not familiar with, like `seq()` below. To investigate what this function does, place your cursor after 'seq' but before the first parenthesis, and press tab. Hover over the function `seq` in the dropdown list to see a full description. If you need more information, you might use the `help()` function, with the name of the function you are curious about inside the parenthesis. The
description indicates that the function is used to generate regular sequences. The information that emerges from using the `help()` function describes a list of arguments, including from, to, and by. This pane will populate on the bottom right side of RStudio when you run the `help()` function. 

```{r seq help}
seq(0, 20, 4)
help(seq)
```


Other ways to get help:

```{r}
?mean # Get help of a particular function
help.search('weighted mean') # search help for a word or phrase.
help(package = 'dplyr') # Get help with package
str(iris) # Get summary of structure of object
class(iris) # find the class an object belongs to.
```

## Objects

You may have heard the phrase "object-oriented programming." This phrase
is accurate as all coding in R relies on the assignment of objects. When
you assign an object in R, you are indicating that you want R to
remember this assignment so it can be used as part of other code. `pi`
is an example of an object built into base R. The input below is not
numeric, but still represents a number. Run the code, and you will see
that the word pi has been assigned the numeric value of pi. This is one
of the few predefined objects in R.

```{r pi object}
pi 
```

You might also assign a number pi to an object you name yourself.

```{r pi assignment}
x <- 3.141593
```

To assign values to objects, as the numerical content of pi was to the
object x, use the `<-` operator. For example, the code segment below
assigns the value 50 to the object 'a', and 14 to the object 'b' using
the `<-` operator. It may be helpful to read the code out loud, saying a
is 50, and b is 14.

```{r twoobjects1}
a <- 50
b <- 14 
```

There are five basic types of objects in R. The objects a, b, and x are
each examples of **vectors**, or homogeneous data made up of characters,
logical, numerical, or integer values. Additionally, a **list** of
heterogeneous data (i.e., involving multiple data types) can be assigned
to an object. **Matricies**, or two-dimensional data with undefined
column headers, and **data frames**, a matrix with multiple specified
columns that represent a certain type of observation seen in
corresponding rows, can also be assigned to objects. Moreover, many
objects are typically involved in coding and they tend to interact with
each other.

See the code chunk below for a simple example of the interaction of two
objects that were assigned in the previous code chunk.

```{r twoobjects2}
a + b
```

Notice that R held the object assignments from the previous cell. You
can also assign a function to an object, and call that object to execute
the function. For instance:

```{r functionobject}
addvalues <- a + b
addvalues
```

The product is not reached because R understands the input `addvalues`,
but because the object add values calls the newly defined function 'a +
b'. Try switching the values of a and b three chunks ago, and running
the subsequent chunks. Remember that objects are case sensitive and
cannot contain spaces. If you ran the code 'A+B', what would happen?

# Case Study

You are approached by a colleague who wants create code that sums
numerical values associated with letters in others' names (e.g., a = 1,
b = 2,... z = 26). To start on this project, create 26 objects, one for
each letter of the alphabet. Then sum your own name using mathematical
expressions and the objects you created.

```{r}
a <- 1
b <- 2
c <- 3 # and so on
```

# Working with Data

## Matricies

Matrices include rectangular data of a single format (numerical or
text). Let's construct a matrix to work with the concept. In the example
below, a sequence is created from 1 to 100, separated by 5s. Two
matricies are then created with the argument for data set to the object
'numbers'. However, the shaping arguments are different for each matrix.

```{r matricies}
numbers <- seq(1, 100, 5)
twocolumns <- matrix(data = numbers, ncol = 5) 
fourrows <- matrix(data = numbers, nrow = 10) 
```

Use the empty box below to call each matrix object that you created.
Notice that when you set the argument 'ncol' the data are shaped based
on the desired number of columns, which is assigned as an argument for
the matrix function.

```{r call}
twocolumns
fourrows

```

## Dataframes

A dataframe contains rectangular with rows and columns, and may contain
different data types. Base R includes several datasets that can be
called into dataframes, and assigned to objects. The View function
allows for matricies and dataframes to be easily accessed by the user.
To view the dataframe 'mtcars', run the code chunk below.

```{r mtcars}
View(mtcars)
```

Try it without the View() function.

```{r mtcars2}
mtcars
```

The head() and tail() functions show the first and last six rows of a
dataframe. Notice that when you run the entire cell, R studio allows you
to scroll through the various outputs.

```{r mtcars3}
head(mtcars)
tail(mtcars)
```

To investigate the names of the columns, run the function colnames().
For a summary of each column of data, call the summary() function.

```{r mtcars4}
colnames(mtcars)
summary(mtcars)
```

Now change the name of column 3 from "disp" to "displace"

```{r mtcars5}
colnames(mtcars)[3] <- "displace"
colnames(mtcars)
```

Using the above code chunk as a model, rename the mpg column to Miles
Per Gallon. Hint: the [3] represents the third column in the dataset.

```{r mtcars6}
colnames(mtcars)[1] <- "Miles Per Gallon"
summary(mtcars)
mtcars$cylinder <- mtcars$cyl
rename(mtcars, horsepower = hp)
```

## Importing and Exporting Data

We will be using a built-in dataset from the `fivethirtyeight` package
that contains information on US Births from 1994 to 2003. Call the
dataset with the object named `'US_births_1994_2003` and assign it to
the object `data` with the `<-` operator. This code can be read as
"create a new object 'data' identical to the dataset
'US_births_1994_2003'." The purpose of renaming this data object is to
simplify subsequent coding practices. That said, you can choose to
rename your dataset anything you'd like, or keep the name as it was
originally stated.

### R-based Data

```{r echo = TRUE, message = F, warning = F}
library(fivethirtyeight)
data <- US_births_1994_2003
```

Practice calling another dataset. Try calling the `data()`function or
visiting the `fivethirtyeight` website and select a dataset that you are
interested in examining. Once you have chosen a dataset, you can assign
that data object to another name, or simply type in the name of the
dataset into a code chunk. Alternatively, you can use the `data()`
funciton with the name of the dataset inside the parenthesis to specify
which dataset you are calling in.

If you would like to import your own data, use the `tidyverse`
conventions below. Remember, you must have the `tidyverse` library
installed and called for this function to work.

### CSV Data

```{r, eval = F}
data <- read_csv('filename.csv')
```

Use caution with xlsx file types that contain multiple sheets. CSVs are
utilized more frequently than xlsx files as `tidyverse` handles them
quickly and simply. To load an xlsx file, you will need to explicitly
install the "readxl" package. You can then call the specific sheet you
want to load using the `sheet` argument. You will need to assign that
sheet to an object to save that particular sheet for further examination
and manipulation.

### XLSX Data

```{r, eval = F}
install.packages('readxl')
library(readxl)
data <- read_excel(spreadsheet, sheet = "sheet1")
```

`Haven` is a package that is associated with, but not inherently a part
of, the `tidyverse` package for SPSS, SAS, and STATA filetypes. You will
need to follow similar conventions as with the readxl library. You can
either define the path using an object, or explictly state the path in
the ultimate function's parenthesis.

### SPSS ".sav" Data

```{r, eval = F}
install.packages('haven')
library(haven)
path <- file.path("example.sav")
data <- read_sav(path)

```

### SAS ".sas7bdat" or ".sas7bcat" Data

```{r, eval = F}
dataset <- read_sas("example.sas7bdat")
```

### For Stata ".dta" Data

```{r, eval = F}
data <- read_dta("example.dta")
```

### Exporting Data

You can export data similarly to how you imported data. See the code
chunk below for examples of export functions for each data filetype. It
is important to note that each of these are exported from the data
object, as it is assumed you will have edited that data object at some
point while using R. Each of these functions will save a dataset in your
working directory based on the characteristics of the object "data" at
the time the function was run.

```{r, eval = F}
write_csv(data)
write_excel(data)
write_sav(data)
write_sas(data)
write_dta(data)
```

